---
title: "Project Proposal"
subtitle: "due October 11, 2021 by 11:59 PM "
author: "Maggie Lundberg, Isabella Kjaerulff, and Riya Mohan"
date: ""
output: pdf_document
---

# Load Packages

```{r load-packages, message = FALSE, warning = FALSE}
library(tidyverse)

``` 

# Load Data
[to be inserted upon receiving expanded OIT container]
```{r load-data, message = FALSE, echo = FALSE}
```

# Introduction and Data, including Research Questions
Our data is provided by NEMSIS, the National EMS Information System, and was collected as an effort to help standardize and improve EMS databases. 

Every year, over 16 million patients are treated by EMS. In comparison to White patients, Asian and Hispanic patients were less likely to be evaluated for their pain. White and Black patients were relatively equal in pain assessment frequency, but only 13.9% of Black patients were given pain medication while 20.1% of White patients were given pain medication (Soares et al, 2019). While there are lots of limitations to interpreting EMS treatment of patients, previous findings have shown that race directly leads to a difference of treatment in emergency situations. We wish to dive deeper into this to see if other factors influence patients’ treatment of care. From the data provided by NEMSIS, we ask, what factors influence somebody's process of emergency care in an EMS call? 

To answer our overarching research question, we plan to center our analysis around these questions:
Is there a correlation between drug use and observed quality of care? 
Is there a correlation between race and observed quality of care?
Are there biases present in the EMS quality of care sample size? 
What is the relationship between symptoms and vital sign ranges?
Is race correlated to technicians' assessment of GCS?

Variables of interest include Primary Symptom, Other Associated Symptoms, Cause of Injury, Barriers to Patient Care, Alcohol/Drug Use Indicators, Protocols Used, Medication Given, Procedure, Procedure Successful/Response to Procedure, Suspected EMS Work Related Exposure/Injury/Death, Race, Initial Patient Acuity vs Final Patient Acuity, and Vital Signs--Blood Pressure, Pulse Oximetry, Respiratory Rate, O2 Saturation, Blood Glucose, Glasgow Coma Score, and Level of Responsiveness.


# Glimpse

```{r glimpsedata}

```

We have achieved extraction of three of our variables with the help of Phuc Nguyen and will continue to upload the full data to Github when it is separated into unidentifiable random samples, either by means of working with an experienced R user or obtaining a larger container from the OIT.


We will be using ~500 observations with 10-20 variables as personally approved by the NEMSIS research team for uploading to Github. We will be including a substantial credit line and clarifying the nature of the data as NOT population-based and NOT used to represent public health data. The process will include uploading the data onto a Windows computer, bypassing the encryption with the corresponding Bitlocker passcode, uploading the zip file onto a high-capacity R-studio container, reading it from SAS to R, joining the files, narrowing down the sample to 500 random observations using R code, saving that file, and uploading it to Github in place of the original data set to work from. Currently, we have an open help desk request of number RITM4124724 to obtain access to the high-capacity container.


# Data Analysis Plan

As a part of our data analysis, a few of the specific predictors we are looking at are age group, gender, disease type, and type of insurance. We will be looking at how these predictors impact United States healthcare spending by the Emergency Department. 

In addition, we have identified a few vital statistical methods needed for analysis.When looking at gender, which we plan to prephase that it is only Male/Female only because of the nature of the dataset, we ask the question, is there a correlation for Males vs. Females? To test this, we will use a t-test for means of males vs. females and perform a linear Regression. Next we ask, is there correlation for males vs. females GIVEN disease type. We plan to perform a regression for this with an interaction term and conditional probability tests. Age, which is a categorical variable, is the next predictor we want to analyze, asking, is there a correlation for various age groups? An f-test for means of age groups using ANOVA, and a step-down tests for the age groups will be done. We will also see if there a correlation for various age groups GIVEN disease type. Additionally, we want to look at the interaction between gender and age, using a linear regression (Main Effects), linear regression (Interaction Effects), and adjusted R^2 for Gender and Age to determine whether the interaction term is useful in the model. Finally we ask, how does the type of insurance play into this? To test this we will use a t-test and linear regression to see how government spending differs by insurance type. 

Visualizations will include linear regression line plots, plots of residuals, and a boxplot of the distribution of spending faceted by insurance type. 

WHAT results are we expecting?



References:



