---
title: "An Investigation of Social Factors Influencing <br> Emergency Healthcare Expenditure"
author: "Maggie Lundberg, Riya Mohan, Izzy Kjaerulff"
institute: "Duke University"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      ratio: "16:9"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r load-packages, include = FALSE}
# Add any additional packages you need to this chunk
# Remove any packages from this list that you're not using
library(tidyverse)
library(tidymodels)
library(palmerpenguins)
library(knitr)
library(xaringanthemer)
```

```{r setup, include=FALSE}
# For better figure resolution
knitr::opts_chunk$set(
  fig.retina = 3, 
  dpi = 300, 
  fig.width = 6, 
  fig.asp = 0.618, 
  out.width = "70%"
  )
```

```{r load-data, include=FALSE}
unzip("../data/IHME_DEX_ED_SPENDING_2006_2016_DATA.zip", exdir =  "../data/spending_data_unzip")

spending <- read.csv("../data/spending_data_unzip/IHME_DEX_ED_SPENDING_2006_2016_DATA_Y2021M09D23.CSV")

spending_malefemale <- spending %>%
  filter(sex %in% c("Female", "Male")) %>%
  filter(age_group_name != "All Ages")

spending_malefemale <- spending_malefemale %>%
  filter(mean_all != 0) %>%
  filter(mean_pub != 0) %>%
  filter(mean_pri != 0) %>%
  filter(mean_oop != 0) %>%
  mutate(lmean_all = log(mean_all)) %>%
  mutate(lmean_pub = log(mean_pub)) %>%
  mutate(lmean_pri = log(mean_pri)) %>%
  mutate(lmean_oop = log(mean_oop))
```

```{r style-slides, echo = FALSE}
style_xaringan(
  title_slide_background_image = "img/markus-spiske-iar-afB0QQw-unsplash.jpg"
)
```

class: inverse, middle, center

background-color: #217718

# Background Information and Significance

Emergency services ensure that individuals can receive timely care for unexpected ailments and injuries, making them a vital component of the healthcare industry. However, in recent years, emergency service spending has seen a significant increase (Scott and Liu 2021), which begs the question of whether access to these service remain equitable to all. Expenditure is one of many ways to investigate healthcare access, which is why we look to understand how various social factors, such as gender and age, affect spending for the US healthcare system's different payers.


---

## Introducing the Data Set

- Our data is provided by the Institute of Health Metrics and Evaluation as part of the Disease Expenditure Project (DEX).

- The health spending data include estimates for U.S. spending on health care divided into three types of payers: public insurance (including Medicare, Medicaid, and other government programs), private insurance, and out-of-pocket payments

- A few of the important variables that will be used in the analysis include: age, gender, disease type, total mean spending.



---
class: inverse, middle, center

background-color: #217718

# Data Analysis

---

## Gender
--

- We performed a t-test looking at the overall differences in log mean emergency department spending between males and females. This t-test showed that for mean spending of all emergency services payment types, the p value of 0.1543 (95% CI -0.0315862, 0.1996079), indicating there is not a significant difference between male and female spending for mean emergency spending.

--

- This t-test showed that for mean spending of all emergency services payment types, the p value is 0.1543 (95% CI -0.0315862, 0.1996079).

--

- Therefore, there is not a significant difference between male and female spending for mean emergency services.

---

## Age
--

- The age variable is split into 19 groups that generally increase by 5-year increments. 
--

-We first used an overall test with ANOVA on the log scale to evaluate the null hypothesis that all of the means for age groups across the years are equal. 
--

- In our F-test (ndf = 18, ddf = 6031), a significant difference among age groups was identified. 

---

## Age
--

- The pairwise t-tests evaluated 99/171 differing combinations, which was consistent with our hypothesis that (a majority of) age group pairs differ in terms of mean expenditures.
--
```{r age-expenditure-boxplot, echo = FALSE, warning = FALSE, message = FALSE, fig.height = 2.5, fig.width = 6}
  
ggplot(data = spending_malefemale, aes(x = age_group_name, y = lmean_all)) +
  geom_boxplot() +
  theme(
    axis.text.x = element_text(color = "grey20", size = 5, 
                               face = "italic", angle = 45,hjust = 1),
    axis.text.y = element_text(color = "grey20", size = 5, face = "italic"),
    axis.title.y = element_text(color = "grey20", size = 8, face = "bold"),
    axis.title.x = element_text(color = "grey20", size = 8, face = "bold"),
    plot.title = element_text(color = "grey20", face = "bold")
        ) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    x = "Age Group",
    y = "Mean Spending, All Payers (ln)",
    title = "Emergency Department General Expenditures"
       )
```


---

## Health Category

---

## Health Category and Gender

---

## Health Category and Age

---

## Conclusion and Discussion

```{r boring-regression}
# a boring regression
model <- lm(dist ~ speed, data = cars)
tidy(model)
glance(model)
```

---

## Acknowledgements

---

## References

Institute for Health Metrics and Evaluation (IHME). United States Healthcare Spending in Emergency Departments byHealth Condition 2006-2016. Seattle, United States of America: Institute for Health Metrics and Evaluation (IHME),2021.

Woody Scott K, Liu A, Chen C, Kaldjian AS, Sabbatini AK, Duber, HC, Dieleman JL. Healthcare Spending in U.S.Emergency Departments by Health Condition, 2006-2016. PLOS One. 27 October 2021.
